---
title: "Differential Expression Results"
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h

output:
  pdf_document:
    fig_caption: true
    keep_tex: true
    latex_engine: pdflatex
  html_document:
    df_print: paged
urlcolor: blue
---


```{r setup3, include=FALSE}
source("config/deseq2_config.R")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.extra = '')
#knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_knit$set(root.dir = rootdir)
```

# DESeq2 Summary Report (with hisat2 mapping) {#deseq2}

This analysis uses DESeq2 to define differentially regulated genes across conditions. The first section deals
with DESeq2 quality assessments and the later sections with differentially expressed genes.

## Number of genes detected

This section provides a look at the number of genes detected i.e read count >=1

```{r detected-genes, fig.cap='Number of genes detected in each sample', fig.height=2.5, fig.width=5, echo=FALSE, message=FALSE}

# get helper functions
source(helper)

# knitr for tables
library(knitr)

mat <- read.csv(feature.counts, header=T, stringsAsFactors=F, sep="\t", row.names=1)
colnames(mat) <- gsub("hisat", "", colnames(mat))
ngenes <- data.frame(colSums(mat >= 1))

# make conditions
ngenes$cond <- makeConds(rownames(ngenes))
colnames(ngenes) <- c("ngenes", "cond")

# plot number of genes
ggplot(ngenes, aes(x=cond, y=ngenes, colour=cond)) + geom_boxplot() + geom_jitter() + theme_bw()

```

### ANOVA of number of genes detected

This is a test of significant differences in the number of genes detected.

```{r sig-difference-in-genes-detected}

summary(aov(ngenes$ngenes~ngenes$cond))

```

\newpage

## Quality control of DESeq2 normalisation procedures

```{r diff-expression, fig.cap='Mean-variance and dispersion estimates from DESeq2', fig.height=4, fig.width=8,  echo=FALSE, message=FALSE}

library(DESeq2)
library(vsn)
library(RSQLite)
library(grid)
library(gridBase)

####################
# analysis
####################

mat <- read.csv(feature.counts, header=T, stringsAsFactors=F, sep="\t", row.names=1)

# hisat suffix remains in colnames so remove this here
colnames(mat) <- gsub(".hisat", "", colnames(mat))

# add condition as coldata
coldata <- makeColData(mat)

# add donor to coldata
donor <- unlist(strsplit(as.character(coldata$sample), "*\\."))
donor <- donor[seq(3, length(donor), 3)]
coldata$donor <- donor

# paired data so remove effect of donor in analysis
dds <- DESeqDataSetFromMatrix(countData = mat,
                              colData = coldata,
			      design = ~ donor + condition)

# add features (gene names)
featureData <- makeFeatureData(annotations.db, mat)
mcols(dds) <- DataFrame(mcols(dds), featureData)

# filtering > a in at least k samples (i.e. all of one condition)
keep <- rowSums(counts(dds) >= a) >= k
dds <- dds[keep,]

# run analysis with donor information as reduced model
dds.lrt <- DESeq(dds, test="LRT", reduced=~donor)
res <- results(dds.lrt)

# get transformed counts
rld <- rlog(dds, blind=TRUE)
rldf <- data.frame(assay(rld))
rldf$gene_id <- rownames(rldf)

# write table out
write.table(rldf, file="genes_rlog.tsv", row.names=F, quote=F, sep="\t")

rld <- read.csv("genes_rlog.tsv", sep="\t", stringsAsFactors=F, header=T)
rownames(rld) <- rld$gene_id
rld <- rld[,c(1:ncol(rld)-1)]


# plot mean-variance relationship and dispersion
# estimates

par(mfrow=c(1,2))
plotMeanSd(rld)
plotDispEsts(dds.lrt)

```

\newpage

# Principle components analysis

Here we perform principle components analysis on normalised read counts (rlog) from DESeq2. This gives an
idea of the structure of the data and whether there is an obvious difference between groups.

```{r pca, fig.cap='Principle components analysis', fig.height=15, fig.width=20,  echo=FALSE, message=FALSE}

df <- rld
pc <- PCA(df)
pcs <- getPCA(pc)

# plot cumulative proportion
p1 <- plotCumulativeProportion(pc)

# plot principle components
p2 <- plotPCA(pc, pcs=c("PC1", "PC2"))
p3 <- plotPCA(pc, pcs=c("PC1", "PC3"))
p4 <- plotPCA(pc, pcs=c("PC2", "PC3"))

layout = rbind(c(1,1,1), c(2,3,4))
grid.arrange(grobs=list(p1,p2,p3,p4), layout_matrix=layout)

```
PC1 clearly reflects the impact of infection and PC2 the variation between donors. It seems clear from the PCA
in this experiment that PC3 represents differences between CPZ and control samples (regardless of infection status).

If we do the PCA again but this time removing the effect of donor then we should see the variation due to CPZ
treatment fall as PC2.

```{r pca.donor, fig.cap="PCA with effect of donor removed", fig.height=5, fig.width=15, message=FALSE, echo=FALSE}

# use limma to remove the batch effect
library(limma)
rld.removed <- data.frame(removeBatchEffect(rld, coldata$donor))

pc.removed <- PCA(rld.removed)
pcs.removed <- getPCA(pc.removed)

# plot principle components
p5 <- plotPCA(pc.removed, pcs=c("PC1", "PC2"))
p6 <- plotPCA(pc.removed, pcs=c("PC1", "PC3"))
p7 <- plotPCA(pc.removed, pcs=c("PC2", "PC3"))

grid.arrange(p5,p6,p7, nrow=1, ncol=3)

```

# Overall differential expression analysis

Here we describe genes that were called as differentially expressed. If there are multiple levels per factor
then it is an overall test (ANOVA-like) whereas if just two conditions are present then it is the direct contrast.
The heatmap produced is using data where the effect of donor has been removed from the expression values.

```{r de, fig.cap='Differential expression analysis', fig.height=15, fig.width=10,  echo=FALSE, message=FALSE}

library(gplots)

res2 <- getResultsTable(res, featureData)

# subset for genes that are differentially regulated
res.diff <- res2[res2$padj < 0.05 & !(is.na(res2$padj)) & !(is.na(res2$log2FoldChange)),]

# get number different
ndiff <- nrow(res.diff)
df <- data.frame("Number DE" = ndiff)
kable(df, caption="number of genes differentially expressed at p < 0.05")

# get differentially expressed gene ids
diffgenes <- res.diff$gene_id
diff <- rld.removed[diffgenes, ]

# heatmap differences
heatmapMatrix(diff)

# table the top 50
toshow <- res.diff[order(res.diff$padj, decreasing=F),][1:50,]
kable(toshow, caption="top 50 differentially expressed genes by adjusted pvalue", row.names=F)

```

# Pattern of regulation in the top genes

As the analysis was done ANOVA style we don't clearly know (without specifying the contrasts) the pattern of regulation
of these genes. Here we plot the top (20) genes to see what the pattern is. The expression values again after the
removal of the effect of donor.

```{r goi, fig.cap='Expression pattern in top 20 genes', fig.height=15, fig.width=15,  echo=FALSE, message=FALSE}

# get gene names
rld2 <- rld.removed

# get rid of any weird duplicates
res.diff <- res.diff[!(duplicated(res.diff$gene_name)),]
res.diff <- res.diff[order(res.diff$padj, decreasing=F),]

# make gene names the rownames
rld2 <- rld2[rownames(res.diff),]
rownames(rld2) <- res.diff[rownames(rld2),]$gene_name

# collect grobs for figure
grob.list <- list()
for (i in 1:20){
    goi <- rownames(rld2)[i]
    p <- plotGOI(rld2, goi=goi)
    grob.list[[i]] <- p
    }

grid.arrange(grobs=grob.list, ncol=4, nrow=5)

```
\newpage

# Specific contrasts

Contrasts of interest in this experiment are Salmonella vs. Control (CPZ+ and CPZ-) as well as CPZ vs Control (Salmonella+ and Salmonella-).
Here we perform these specific contrasts to get an idea of differentially expressed genes.

```{r contrasts, fig.cap='MA plot for each contrast', fig.height=8, fig.width=14,  echo=FALSE, message=FALSE}

#########################################
# get contrasts (Salmonella vs. Control)
#########################################

dds.sal.control <- dds[,1:8]
dds.sal.control$condition <- droplevels(dds.sal.control$condition)
dds.sal.control$condition <- factor(dds.sal.control$condition, levels=c("Control", "Salmonella"))
dds.sal.control <- DESeq(dds.sal.control, test="LRT", reduced=~donor)
res.sal.control <- results(dds.sal.control)

##############################
# Sal + CPZ vs CPZ 
##############################

dds.salcpz.cpz <- dds[,9:16]
dds.salcpz.cpz$condition <- droplevels(dds.salcpz.cpz$condition)
dds.salcpz.cpz$condition <- factor(dds.salcpz.cpz$condition, levels=c("cpz", "cpzSalmonella"))
dds.salcpz.cpz <- DESeq(dds.salcpz.cpz, test="LRT", reduced=~donor)
res.salcpz.cpz <- results(dds.salcpz.cpz)


##############################
# Sal + CPZ vs control 
##############################

dds.salcpz.control <- dds[,c(1:4, 13:16)]
dds.salcpz.control$condition <- droplevels(dds.salcpz.control$condition)
dds.salcpz.control$condition <- factor(dds.salcpz.control$condition, levels=c("Control", "cpzSalmonella"))
dds.salcpz.control <- DESeq(dds.salcpz.control, test="LRT", reduced=~donor)
res.salcpz.control <- results(dds.salcpz.control)

#############################
# CPZ vs. Control
#############################

dds.cpz.control <- dds[,c(1:4,9:12)]
dds.cpz.control$condition <- droplevels(dds.cpz.control$condition)
dds.cpz.control$condition <- factor(dds.cpz.control$condition, levels=c("Control", "cpz"))
dds.cpz.control <- DESeq(dds.cpz.control, test="LRT", reduced=~donor)
res.cpz.control <- results(dds.cpz.control)

#############################
# CPZSal vs. Sal
#############################

dds.salcpz.sal <- dds[,c(5:8,13:16)]
dds.salcpz.sal$condition <- droplevels(dds.salcpz.sal$condition)
dds.salcpz.sal$condition <- factor(dds.salcpz.sal$condition, levels=c("Salmonella", "cpzSalmonella"))
dds.salcpz.sal <- DESeq(dds.salcpz.sal, test="LRT", reduced=~donor)
res.salcpz.sal <- results(dds.salcpz.sal)


###############################
# compare fold changes between
# both salmonella (+/-CPZ) vs
# control to see enhancement
# of salmonella response
###############################

dat <- data.frame(sal.control = res.sal.control$log2FoldChange,
                  salcpz.control = res.salcpz.control$log2FoldChange,
		  cpz.control = res.cpz.control$log2FoldChange,
		  salcpz.cpz = res.salcpz.cpz$log2FoldChange,
		  salcpz.sal = res.salcpz.sal$log2FoldChange)
dat$significant <- ifelse(res.sal.control$padj < 0.05, "yes", "no")

cor1 <- cor(dat$sal.control, dat$salcpz.control, use="pairwise.complete.obs")

p1 <- ggplot(dat, aes(x=sal.control, y=salcpz.control, colour=significant))
p2 <- p1 + geom_point(pch=18, size=0.5)
p3 <- p2 + xlab("log2(Fold change) Salmonella vs. Control")
p4 <- p3 + ylab("log2(Fold change) Salmonella + CPZ vs. Control")
p5 <- p4 + theme_bw()
p6 <- p5 + geom_abline(intercept=0, slope=1)
p7 <- p6 + geom_abline(intercept=1, slope=1, linetype="dashed")
p8 <- p7 + geom_abline(intercept=-1, slope=1, linetype="dashed")
p9 <- p8 + scale_colour_manual(values=c("grey", "red4"))
p10 <- p9 + annotate(geom="text", label=paste("r=", round(cor1,2), sep=""), x=5, y=15)
ggsave("scatterplot_salmonella_foldchanges.pdf")


###
# fold change comparison between Salmonella vs. Control and CPZ vs. control
###

cor2 <- cor(dat$sal.control, dat$cpz.control, use="pairwise.complete.obs")
p1 <- ggplot(dat, aes(x=sal.control, y=cpz.control, colour=significant))
p2 <- p1 + geom_point(pch=18, size=0.5)
p3 <- p2 + xlab("log2(Fold change) Salmonella vs. Control")
p4 <- p3 + ylab("log2(Fold change) CPZ vs. Control")
p5 <- p4 + theme_bw()
p6 <- p5 + geom_smooth(data=dat, aes(x=salcpz.cpz, y=salcpz.sal),method="lm", inherit.aes=F, colour="black")
p7 <- p6 + scale_colour_manual(values=c("grey", "red4"))
p8 <- p7 + annotate(geom="text", label=paste("r=", round(cor2,2), sep=""), x=5, y=15)
ggsave("scatterplot_cpz_foldchanges.pdf")


###
# fold change comparison between Salmonella + CPZ vs. CPZ and Salmonella + CPZ vs. Salmonella
###

cor3 <- cor(dat$salcpz.cpz, dat$salcpz.sal, use="pairwise.complete.obs")
p1 <- ggplot(dat, aes(x=salcpz.cpz, y=salcpz.sal, colour=significant))
p2 <- p1 + geom_point(pch=18, size=0.5)
p3 <- p2 + xlab("log2(Fold change) Salmonella + CPZ vs. CPZ")
p4 <- p3 + ylab("log2(Fold change) Salmonella + CPZ vs. Salmonella")
p5 <- p4 + theme_bw()
p6 <- p5 + geom_smooth(data=dat, aes(x=salcpz.cpz, y=salcpz.sal),method="lm", inherit.aes=F, colour="black")
p7 <- p6 + scale_colour_manual(values=c("grey", "red4"))
p8 <- p7 + annotate(geom="text", label=paste("r=", round(cor3,2), sep=""), x=5, y=15)
ggsave("scatterplot_salmonellacpz_foldchanges.pdf")

####################
# full set from LRT
####################

lrt.set <- diffgenes

p1 <- MAPlot(res.sal.control, lfc=lfc, test.in=T, test.set=lrt.set, title="Sal vs. Control")
ggsave("sal.control.ma.pdf")
p2 <- MAPlot(res.salcpz.cpz, lfc=lfc, test.in=T, test.set=lrt.set, title="Sal + CPZ vs. CPZ")
ggsave("salcpz.cpz.ma.pdf")
p3 <- MAPlot(res.cpz.control, lfc=lfc, test.in=T, test.set=lrt.set, title="CPZ vs. Control")
ggsave("cpz.control.ma.pdf")
p4 <- MAPlot(res.salcpz.sal, lfc=lfc, test.in=T, test.set=lrt.set, title="Sal + CPZ vs. Sal")
ggsave("salcpz.sal.ma.pdf")

grid.arrange(p1,p2,p3,p4, nrow=2, ncol=2)

# write out lists
sal.control.diff <- getResultsTable(res.sal.control, featureData)
sal.control.diff <- sal.control.diff[sal.control.diff$padj < 0.05 & abs(sal.control.diff$log2FoldChange) > lfc & !(is.na(sal.control.diff$padj)) & !(is.na(sal.control.diff$log2FoldChange)) & sal.control.diff$gene_id %in% lrt.set, ]

salcpz.cpz.diff <- getResultsTable(res.salcpz.cpz, featureData)
salcpz.cpz.diff <- salcpz.cpz.diff[salcpz.cpz.diff$padj < 0.05 & abs(salcpz.cpz.diff$log2FoldChange) > lfc & !(is.na(salcpz.cpz.diff$padj)) & !(is.na(salcpz.cpz.diff$log2FoldChange)) & salcpz.cpz.diff$gene_id %in% lrt.set, ]

cpz.control.diff <- getResultsTable(res.cpz.control, featureData)
cpz.control.diff <- cpz.control.diff[cpz.control.diff$padj < 0.05 & abs(cpz.control.diff$log2FoldChange) > lfc & !(is.na(cpz.control.diff$padj)) & !(is.na(cpz.control.diff$log2FoldChange)) & cpz.control.diff$gene_id %in% lrt.set, ]

salcpz.sal.diff <- getResultsTable(res.salcpz.sal, featureData)
salcpz.sal.diff <- salcpz.sal.diff[salcpz.sal.diff$padj < 0.05 & abs(salcpz.sal.diff$log2FoldChange) > lfc & !(is.na(salcpz.sal.diff$padj)) & !(is.na(salcpz.sal.diff$log2FoldChange)) & salcpz.sal.diff$gene_id %in% lrt.set, ]

sal.control.up <- sal.control.diff$gene_name[sal.control.diff$log2FoldChange > lfc]
sal.control.dn <- sal.control.diff$gene_name[sal.control.diff$log2FoldChange < (-lfc)]

salcpz.cpz.up <- salcpz.cpz.diff$gene_name[salcpz.cpz.diff$log2FoldChange > lfc]
salcpz.cpz.dn <- salcpz.cpz.diff$gene_name[salcpz.cpz.diff$log2FoldChange < (-lfc)]

cpz.control.up <- cpz.control.diff$gene_name[cpz.control.diff$log2FoldChange > lfc]
cpz.control.dn <- cpz.control.diff$gene_name[cpz.control.diff$log2FoldChange < (-lfc)]

salcpz.sal.up <- salcpz.sal.diff$gene_name[salcpz.sal.diff$log2FoldChange > lfc]
salcpz.sal.dn <- salcpz.sal.diff$gene_name[salcpz.sal.diff$log2FoldChange < (-lfc)]

######
# write

write.table(sal.control.up, file="../data/sal.control.up.tsv", sep="\t", row.names=F, col.names=F, quote=F)
write.table(sal.control.dn, file="../data/sal.control.dn.tsv", sep="\t", row.names=F, col.names=F, quote=F)

write.table(salcpz.cpz.up, file="../data/salcpz.cpz.up.tsv", sep="\t", row.names=F, col.names=F, quote=F)
write.table(salcpz.cpz.dn, file="../data/salcpz.cpz.dn.tsv", sep="\t", row.names=F, col.names=F, quote=F)

write.table(cpz.control.up, file="../data/cpz.control.up.tsv", sep="\t", row.names=F, col.names=F, quote=F)
write.table(cpz.control.dn, file="../data/cpz.control.dn.tsv", sep="\t", row.names=F, col.names=F, quote=F)

write.table(salcpz.sal.up, file="../data/salcpz.sal.up.tsv", sep="\t", row.names=F, col.names=F, quote=F)
write.table(salcpz.sal.dn, file="../data/salcpz.sal.dn.tsv", sep="\t", row.names=F, col.names=F, quote=F)

# background file
features <- makeFeatureData(annotations.db, rld)
write.table(as.character(features$gene_name), file="../data/background.tsv", sep="\t", col.names=F, row.names=F, quote=F)

###########################################
# write out differentially expressed data
###########################################

write.table(sal.control.diff, file="../data/sal_vs_control.diff", sep="\t", row.names=F, quote=F)
write.table(salcpz.cpz.diff, file="../data/salcpz_vs_cpz.diff", sep="\t", row.names=F, quote=F)
write.table(cpz.control.diff, file="../data/cpz_vs_control.diff", sep="\t", row.names=F, quote=F)
write.table(salcpz.sal.diff, file="../data/salcpz_vs_sal.diff", sep="\t", row.names=F, quote=F)

```

\newpage

# Genes of interest

Given the structure of the data and the increased ability of CPZ-treated macrophages to kill Salmonella it is of
interest to see if CPZ upregulates genes that are part of the classic Salmonella response. To look at this
I have taken genes that are up-regulated in the Salmonella vs. Control comparison and intersected this set with
the set of genes up-regulated in CPZ vs. Control. Genes that are not different between CPZ and Control in infection
are kept.

```{r intersection, fig.cap="Salmonella-inducible genes regulated by CPZ in steady state", fig.height=9, fig.width=12, echo=FALSE, message=FALSE}

gois <- intersect(sal.control.up, cpz.control.up)
gois <- setdiff(gois, salcpz.cpz.up)
gois <- setdiff(gois, salcpz.cpz.dn)


kable(data.frame("Total" = length(gois)))

# collect grobs for figure
grob.list <- list()
for (i in 1:length(gois)){
    goi <- gois[i]
    p <- plotGOI(rld2, goi=goi)
    grob.list[[i]] <- p
    }

grid.arrange(grobs=grob.list, ncol=4, nrow=4)


```



